---
title: "Finding and loading datasets"
description: "Datasets are the primary unit of data in Crunch. Here's how to find and load existing datasets in your R session."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[Previous: Getting started](crunch.html)

```{r, results='hide', echo=FALSE, message=FALSE}
library(crunch)
load("vignettes.RData")
options(width=120)
```

To get started working with Crunch data in R, we need to load a dataset in our local R session. This doesn't download all of the data: it pulls some metadata and references to where in the API everything else exists, which other methods will use to make requests as needed.

Following the introductory `vignette("crunch")`, which shows how to import data to Crunch, this document covers how to find and load existing datasets. If you have a dataset you want to put on Crunch, please see `vignette("crunch")`.

# Finding datasets

## listDatasets (including projects)

## searchDatasets

# Loading datasets

## loadDataset by name (and project)

## load with URL (share email, web browser)

* note also `webApp()` to go the other way

<!-- Clean up what follows -->

## Loading existing Crunch datasets

Datasets already existing on the Crunch server can be loaded with `loadDataset`. The function takes either the dataset's name, or the position within the dataset list returned by `listDatasets`:

```{r, eval=FALSE}
listDatasets()
```
```
## [1] "Economist/YouGov Weekly Survey"
```
```{r, eval=FALSE}
ds <- loadDataset("Economist/YouGov Weekly Survey")
```
```{r}
is.dataset(ds)
```

[Next: variable metadata](variables.html)
