---
title: "Crunch Deck Cookbook"
description: "The crunch package makes it easy to save analyses in a scalable way. This cookbook shows you how to set the components of a slide so that you have exactly the analysis you want."
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Crunch Deck Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Decks are a way to save a set of analyses so that you or your team can refer back to them later,
export to Excel or PowerPoint, or create a Crunch Dashboard. Each *deck* is made up of a set of
*slides*, and each *slide* contains an *analysis* and optionally a *title* and *subtitle*.

While a good slide generally appears simple to the viewer, a lot depends on getting the analysis
exactly right, and so creating them does require setting the analysis's attributes just right. The
attributes of a slide's analysis are:

- **a query**: A set of variables and what calculation to perform (such as a count or mean)
- **the query environment**: Modifiers to the data that the query is performed on, like a filter
or weight to use.
- **dimension transformations**: A dimension is a set of items that the calculation
is enumerated along (a table based on a categorical array could have the categories in the
"rows" dimension and the subvariables in the "columns" dimension). Dimension transformations are
modifications to how those dimensions are displayed, such as reordering, renaming or assigning
colors.
- **display settings and visual specifications**: There's a whole other collection of settings
such as the type of visualization, number of decimals, and advanced options for how the slide
is exported to Excel.

The recipes in this cookbook all start from the pets example dataset (available from 
`newExampleDataset()`).

```{r}
suppressPackageStartupMessages({
    library(purrr)
    library(crunch)
})

options("crunch.show.progress" = FALSE)

login()
ds <- newExampleDataset()
```

## Basic deck and slide manipulation

### Creating a slide on a new deck
#### Problem
You want to create a new deck and add slides to it.

#### Solution
A deck is created on a dataset with the command `newDeck()`. It takes the dataset, a title for the
deck, and can take `is_public=TRUE` if the deck should be made public to other users of the dataset
(it defaults to `FALSE`).

After the deck is created, the `newSlide()` function adds a slide.

```{r}
deck <- newDeck(ds, "Q3 Pets Deck", is_public = TRUE)
private_deck <- newDeck(ds, "Private Deck")

# If no `vizType` is specified, defaults to a table
slide <- newSlide(deck, ~q1, title = "Table of Favorite Pet")

# Example of setting a vizType and filter
slide <- newSlide(
    deck, 
    mean(ndogs) ~ country, 
    title = "Dot Plot of Mean Dogs by Country",
    display_settings = list(vizType = "dotplot"),
    filter = ds$q1 == "Dog"
)

deck <- refresh(deck)
```


### Creating a slide on an existing deck
#### Problem
You want to add a slide to a deck that has already been made.

#### Solution
The `decks()` function access the decks catalog for a dataset. You can select one
name or position and then add to it.

```{r}
ds <- refresh(ds)
decks(ds)

private_deck <- decks(ds)[["Private Deck"]]

slide <- newSlide(
    private_deck, 
    ~q1, 
    title = "Bar Plot of Favorite Pet", 
    display_settings = list(vizType = "groupedBarPlot")
)
```


### Editing an existing slide
#### Problem
You want to edit a slide that's already been created.

#### Solution
Slides can be accessed from the deck's slide catalog, available from the `slides()` command. 
You can retrieve them by their title or position. 

The helper functions `title<-`, `subtitle<-`, `query<-`, `weight<-`, `filter<-`, `transforms<-`,
`displaySettings<-` and `vizSpecs<-` help set options on a slide.

```{r}
# Move title to subtitle and change the title
slide <- slides(deck)[["Table of Favorite Pet"]]
subtitle(slide) <- title(slide)
title(slide) <- "Cats are the most popular"


# Rename a category
slide <- slides(deck)[[2]]
transforms(slide) <- list(
    rows_dimension = makeDimTransform(rename = c("AUS" = "Australia"))
)
```


```{r, include=FALSE}
### Reordering slides on a deck
#### Problem
# You want to rearrange the order slides on an existing deck.

#### Solution
# TODO: This is not currently possible from R :(
# (Note this controls the order of the slides in a deck, which controls how they appear in the web app's
# deck viewer and Excel and PowerPoint exports, but does not change order or position of an existing 
# Crunch Dashboard)
```


### Deleting slides on a deck
#### Problem
You want to delete a slide from a deck.

#### Solution
Access a slide from the slide catalog and then use the `delete()` command to delete it (it will
ask before deleting unless you use command `with_consent()`).

```{r}
slide <- slides(deck)[[1]]

if (FALSE) { # Not actually run for example
    delete(slide)
}
```

### Making a deck public (or private)
#### Problem
You want to change a deck from being private to public (or vice-versa).

#### Solution
The `is.public<-` function can set the deck's status.

```{r}
is.public(private_deck) <- TRUE # now public
```



## Query types
Queries define the variables and summary measures used for the slide's analysis. They use the 
formula notation used by the crunch function `crtabs()` which is based on base R's `xtabs()`.

### Univariate frequencies
#### Problem
You want to get the frequencies of a single categorical or multiple response variable.

#### Solution
The query for a univariate count query puts the variable on the right hand side of a
formula (eg `~var`).

```{r}
slide <- newSlide(
    deck, 
    ~q1, 
    title = "Univariate frequency: Favorite Pet"
)
```

### Bivariate frequencies (crosstabs)
#### Problem
You want to get a crosstab (or a frequency from two variables' joint distribution)

#### Solution
The query for a multivariate frequency uses the `+` to separate the variables on the right
hand side of the formula (eg `~var1 + var2`).

```{r}
slide <- newSlide(
    deck, 
    ~q1 + country, 
    title = "Bivariate frequency: Favorite Pet by country"
)

# A third dimension is possible, which will usually result in a tabbed result:
slide <- newSlide(
    deck, 
    ~q1 + country + wave, 
    title = "Trivariate frequency: Favorite Pet by country by wave"
)
```


### Frequencies from a Categorical Array
#### Problem
You want to get the frequencies from a categorical array variable

#### Solution
A categorical array contributes two dimensions to the analysis, a "categories" dimension
and a "subvariables" dimension. If your query just specifies the variable, by default the 
categories dimension is used first and the categories second, but you can specify the order
by using `categories()` and `subvaribles()` functions in your query.

```{r}
slide <- newSlide(
    deck, 
    ~allpets, 
    title = "Categorical array: default order"
)

slide <- newSlide(
    deck, 
    ~categories(allpets) + subvariables(allpets), 
    title = "Categorical array: categories on rows dimension"
)
```


### Mean of a Numeric variable
#### Problem
You want to get the mean from a Numeric (Numeric Array) variable

#### Solution
A numeric summary measure like a mean goes on the left hand side of the formula in a query.
The right hand side cannot be empty, but to get the mean of the whole dataset put `1`.
```{r}
slide <- newSlide(
    deck, 
    mean(ndogs) ~ 1,
    title = "Mean Number of Dogs"
)

slide <- newSlide(
    deck, 
    mean(ndogs) ~ country,
    title = "Mean Number of Dogs by Country"
)
```


### Scorecard of aligned Multiple Response variables
#### Problem
You want to make comparisons of frequencies of a set of Multiple Response variables with
the same items (response)

#### Solution
A scorecard is a rectangular grid of different Multiple Response variables with their
items aligned. The query for a scorecard can be created using the `scorecard()` function.

```{r}
# There's only one MR available on this dataset, so we repeat the same one twice to illustrate
slide <- newSlide(
    deck, 
    ~scorecard(allpets, allpets), 
    title = "Scorecard"
)
```


## Dimension transforms
Query results have "dimensions", which are enumerated sets that the calculation's results
are formed in, such as the categories of a categorical variables or the items in a 
multiple response variables. Their behavior in the slide can be customized using 
dimension transforms.

A query result generally has up to three dimensions. The first is the "rows_dimension", second
is the "columns" dimension and third is the "tabs_dimension". When using the `transform`
argument of `newSlide()` or setting the `transforms<-` of a slide directly, you form a
named list with these dimensions as the names. The helper function `makeDimTransform()`
can also help create the dimension changes.

### Using a dataset's palette
#### Problem
You want to make the colors of a dashboard tile use a pre-defined palette.

#### Solution
Each Crunch Dataset has a set of color palettes associated with it's account and folder.
You can access the palettes using the `palettes()` or `defaultPalette()` functions.
Then using the `makeDimTransform()` function you can use this palette. The colors are
used in the order they appear and if more colors are needed than provided by the 
palette, the default colors are used.

```{r}
slide <- newSlide(
    deck, 
    ~q1, 
    title = "Favorite pet using default palette",
    display_settings = list(vizType = "groupedBarPlot"),
    transform = list(
        rows_dimension = makeDimTransform(colors = defaultPalette(ds))
    )
)

graph_pal <- palettes(ds)[["purple palette"]]
slide <- newSlide(
    deck, 
    ~categories(petloc) + subvariables(petloc), 
    title = "Pets by location using another palette",
    display_settings = list(vizType = "horizontalBarPlot"),
    transform = list(
        rows_dimension = makeDimTransform(colors = graph_pal)
    )
)
```


### Using a set of colors from R
#### Problem
You want to make the colors of a dashboard tile use a set of colors you specify in the script

#### Solution
If you want to specify the colors manually, you can also use a character vector of RGB hex codes.

```{r}
slide <- newSlide(
    deck, 
    ~q1, 
    title = "Favorite pet using colors from R",
    display_settings = list(vizType = "groupedBarPlot"),
    transform = list(
        rows_dimension = makeDimTransform(colors = c("#af8dc3", "#f7f7f7", "#7fbf7b"))
    )
)
```


### Hiding a dimension item
#### Problem
You want to hide a dimension item (a category or subvariable) from the slide.

#### Solution
The `hide` argument of `makeDimTransform()` takes a category name or id, if 
the dimension is made from categories, or a subvariable name or alias if the
dimension is made from subvariables (as in a Multiple Response variable or a
subvariables dimension of a Categorical Array or Numeric Array).
```{r}
slide <- newSlide(
    deck, 
    ~q1, 
    title = "Favorite pet excluding birds",
    display_settings = list(vizType = "groupedBarPlot"),
    transform = list(
        rows_dimension = makeDimTransform(hide = "Bird")
    )
)
```


## Display settings and Visual Specifications

### Choosing a graph type
#### Problem
You want to create a slide with a display type other than table.

#### Solution
The default display of a tile is the table, but the `vizType` display setting chooses
between other options. The most commonly used vizTypes are: 
- `table` (always available)
- `groupedBarPlot`, `stackedBarPlot`, `horizontalBarPlot`, `horizontalStackedBarPlot` (available
for queries based on a count in any number of dimensions)
- `timeplot` (available when the second dimension has a time component)
- `dotplot` (available for displays of means)
- `donut` (available only for 1 dimensional count queries)

```{r}
slide <- newSlide(
    deck, 
    ~q1 + country, 
    title = "Favorite pet by country horizontal bar plot",
    display_settings = list(vizType = "horizontalBarPlot")
)


slide <- newSlide(
    deck, 
    ~q1 + wave, 
    title = "Favorite pet over time timeplot",
    display_settings = list(vizType = "timeplot")
)
```

### Using a slide as a template
#### Problem
You want to use the settings from an existing slide to create a new one (or modify an existing one).

#### Solution
The functions `displaySettings()` and `vizSpecs()` give access to the settings on an existing slide.
This slide can be a slide you've created from R or from the web app, so that you can use the visual
editor to perfect the look for one slide and then use it for a whole set of slides. You can either 
set the attributes directly, or use `dput()` to print out the object in a way that you can copy
and paste into your code.

```{r}
template_deck <- newDeck(ds, "Templates", is_public = TRUE)
slide <- newSlide(
    template_deck, 
    ~q1, 
    title = "Donut with value labels",
    display_settings = list(vizType = "donut", showValueLabels = TRUE),
    viz_specs = list(
        default = list(
            format = list(
                decimal_places = list(percentages = 0L, other = 2L),
                show_empty = FALSE
            )
        )
    )
)

# Setting the slide `display_setting` and `viz_specs` directly:
slide <- newSlide(
    deck, 
    ~country,
    title = "Country donut with value labels",
    display_settings = displaySettings(template_deck[["Donut with value labels"]]),
    viz_specs = vizSpecs(template_deck[["Donut with value labels"]])
)

# How to print out the structure in a format that can be copy and pasted into your code
print(dput(displaySettings(template_deck[["Donut with value labels"]])))
```



## Bulk Creation of Slides
Sometimes you want to make many slides with related formatting to create a document
that gives a good high level overview of a dataset. The [`tabBook()`] function is designed 
to create a basic "top line" report  of simple crosstabs from a multitable, and is probably
the first thing you should check if you're thinking of making bulk analyses. However, `tabBook()`
does not allow for all of the customization possible in a slide.

The trickiest part of bulk creating slides from R is iterating over the variables. The general
behind all of these cookbook recipes is to get a list of variable aliases, iterate over them 
using them to get other variable metadata. The trickiest part is to create a query formula
from a string, but the `as.formula()` function helps with this. This cookbook uses base R functions
`lapply()` and `paste0()`, but the "tidyverse" functions `purrr::walk()` and `glue::glue()` are 
well-suited to this task.


### Basic Topline for Entire Dataset
#### Problem
You want to create a simple report for every variable in a dataset.

#### Solution
Use the `variables()` function to get the variables from a dataset, and the `aliases()`
function to get their aliases. Then use `lapply()` to iterate over the variable aliases
and construct the slide using `paste0()` and `as.formula()`.
```{r}
deck <- newDeck(ds, "Full Dataset Topline Deck", is_public = TRUE)

var_aliases <- aliases(variables(ds))

slides <- lapply(var_aliases, function(alias) {
    slide_query <- as.formula(paste0("~", alias))
    slide_title <- paste0("Topline - ", name(ds[[alias]]))
    
    newSlide(deck, slide_query, title = slide_title)
})
```

### Slides for Variables Within a Particular Folder
#### Problem
You want to create a simple report for every variable in a particular folder.

#### Solution
The `variables()` function can also work on a folder, so we can make a deck from
variables in a folder in a similar way to making one for a whole dataset.
```{r}
deck <- newDeck(ds, "Folder Topline Deck", is_public = TRUE)

folder <- cd(ds, "Key Pet Indicators")
var_aliases <- aliases(variables(folder))

slides <- lapply(var_aliases, function(alias) {
    slide_query <- as.formula(paste0("~", alias))
    slide_title <- paste0("Topline - ", name(ds[[alias]]))
    
    newSlide(deck, slide_query, title = slide_title)
})
```


#### Problem
You want to create crosstabs for many variables across a set of variables.

#### Solution
You can use `lapply()` to iterate over both the row and column variables of
the crosstab.
```{r}
deck <- newDeck(ds, "Crosstabs Deck", is_public = TRUE)

demo_vars <- aliases(variables(cd(ds, "Dimensions")))
var_aliases <- setdiff(aliases(variables(ds)), demo_vars) # don't cross demo vars with themselves

slides <- lapply(var_aliases, function(alias) {
    # Add a slide before crosstabs of the univariate frequency
    all_query <- as.formula(paste0("~", alias))
    all_title <- paste0("Frequency - ", name(ds[[alias]]))
    
    newSlide(deck, all_query, title = all_title)
    
    lapply(demo_vars, function(demo_alias) {
        crosstab_query <- as.formula(paste0("~", demo_alias, " + ", alias))
        crosstab_title <- paste0("Crosstab - ", name(ds[[alias]]), " by ", name(ds[[demo_alias]]))
        
        newSlide(deck, crosstab_query, title = crosstab_title)
    })
})
```


### Slides Based on Variable Type
#### Problem
You want to create a report with slides that vary based on the variable's type.

#### Solution
You can create functions that create slides for a particular variable type and
then choose which function to use based on the variable's type while iterating.

```{r}
cat_slide <- function(alias, ds, deck) {
    slide_query <- as.formula(paste0("~", alias))
    slide_title <- paste0(name(ds[[alias]]))
    newSlide(
        deck, 
        slide_query,
        title = slide_title,
        display_settings = list(vizType = "donut")
    )
}

mr_slide <- function(alias, ds, deck) {
    slide_query <- as.formula(paste0("~", alias))
    slide_title <- paste0(name(ds[[alias]]))
    newSlide(
        deck, 
        slide_query,
        title = slide_title,
        display_settings = list(vizType = "groupedBarPlot")
    )
}

numeric_slide <- function(alias, ds, deck) {
    slide_query <- as.formula(paste0("mean(", alias, ") ~ wave"))
    slide_title <- paste0(name(ds[[alias]]), " over time")
    newSlide(
        deck, 
        slide_query,
        title = slide_title,
        display_settings = list(vizType = "timeplot")
    )
}

deck <- newDeck(ds, "Slides Customized by Variable Type", is_public = TRUE)

var_aliases <- c("q1", "allpets", "ndogs")
slides <- lapply(var_aliases, function(alias) {
    switch(
        type(ds[[alias]]),
        "categorical" = cat_slide(alias, ds, deck),
        "multiple_response" = mr_slide(alias, ds, deck),
        "numeric" = numeric_slide(alias, ds, deck),
    )
})
```
