---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(crunch.lake)
library(tidyr)
library(dplyr)
library(ggplot2)
library(generics)   # I have a bug in my NAMESPACE I'm still figuring out so this is required for as.factor.crunch_array_variable

crunch::setupCrunchAuth("yougov")
ds <- crunch::loadDataset("https://yougov.crunch.io/api/datasets/58c9c65c01f54b51985abe5fbad6e3d5/")

project_dir <- "~/crunch_lake_vignette"

if (!dir.exists(project_dir)) {
    cr_export(ds, path = project_dir)
}

# You don't have to specify a folder, but this puts data in a temp folder, which is deleted 
# at end of R session.
# data <- cr_read_data(ds)
```

## Metadata

The Crunch Lake metadata is read in a sensible form (though still needs some work)

```{r metadata}
md <- cr_read_meta(project_dir)

md
```

## Data

### Converted to base types (`variable_format = "base"`)
The default behavior is to load variables into base R types. Multiple Response
variables become logical, all other categoricals become factors, and 
numeric/text/datetime become their native R equivalents. While, data in this 
structure can be easier to work with, it does mean that the
only way to refer to the metadata is from the metadata (from `cr_read_meta()`)
and some fidelity is lost (for example you can't distinguish missing categories from
each other).
```{r data}
data <- cr_read_data(project_dir)

data
```


### With crunch metadata (`variable_format = "vctrs"`)
By specifying `variable_format = "vctrs"`, you can read the data with columns
that are R objects with full fidelity of the crunch metadata.
```{r data_vctrs}
data <- cr_read_data(project_dir, variable_format = "vctrs")

data
```


The `crunch_categorical_variable` is an improvement on `haven::labelled()` because it prevents combining
incompatible category types:

```{r categorical}
# Lets you combine categories that exactly align
bind_rows(
    tibble(x = data$enjoy_mr$enjoy_mr_savory),
    tibble(x = data$enjoy_mr$enjoy_mr_spicy)
)

data |>
    mutate(wave = if_else(wave == 1, 2, wave)) |>
    count(wave)


```

```{r, error=TRUE}
# But prevents you from combining incompatible categories (currently any difference is fatal)
bind_rows(
    tibble(x = data$enjoy_mr$enjoy_mr_savory),
    tibble(x = data$wave)
)
```
```{r, error=TRUE}
data |>
    mutate(wave = if_else(wave == 1, 1000, wave)) |>
    count(wave)
```


In many situations, you'll want to convert to base R data types. You can do this manually:
```{r convert}
data |>
    transmute(
        wave_date = as.Date(wave),
        wave = as.factor(wave),
        enjoy_mr = as.logical(enjoy_mr),
        veg_enjoy_ca = as.factor(veg_enjoy_ca),
        ratings_numa = as.numeric(ratings_numa)
    )
```


The metadata can be useful though. Here are some things you may want to do:


Add a categorical variable with their favorite vegetable + numeric with the rating
(Keeping metadata would allow you to send the data back, though this hasn't been worked on yet)

```{r new_var}
data <- data |>
    mutate(
        highest_rated = crunch_categorical_variable(
            apply(ratings_numa, MARGIN = 1, which.max),
            label = "Highest Rated Vegetable",
            path = "/public/Additions/",
            values = tibble::tibble(
                value = seq_len(ncol(ratings_numa)),
                label = purrr::map_chr(ratings_numa, label),
                missing = FALSE, 
                scale = NA,
                selected = NA,
                date = NA
            )
        ),
        highest_rating = crunch_numeric_variable(
            apply(ratings_numa, MARGIN = 1, max, na.rm = TRUE),
            label = "Highest Vegetable Rating",
            path = "/public/Additions/"
        )
    )

```


Use folder structure for dplyr selections
```{r path_manip}
data |> 
    select(
        ratings_numa,
        where(~startsWith(path(.), "/public/Additions") & !is.na(path(.)))
    )
```


Make a graph with axis that has dates and wave labels

```{r time_graph}
time_series <- data |>
    mutate(ratings_numa = rename_with(ratings_numa, ~purrr::map_chr(., ~label(data$ratings_numa[[.]])))) |>
    group_by(wave) |> 
    summarize(mean(ratings_numa, na.rm = TRUE)) |>
    pivot_longer(-wave, names_to = "Vegetable")

ggplot(time_series, aes(x = as.Date(wave), y = value, color = Vegetable)) +
    geom_point() +
    scale_x_date(
        label(data$wave),
        breaks = as.Date(unique(time_series$wave)), 
        labels = as.character(unique(time_series$wave)),
        sec.axis = sec_axis(identity)
    ) + scale_y_continuous(
        label(data$ratings_numa)
    )
```


2 Dimensional arrays (experimental support)
```{r two_dim}
data$funnel_mr_2d <- crunch_categorical_array_variable(
    tibble::tibble(
        aware = data$funnel_aware_mr |> setNames(c("jicama", "kohlrabi")),
        consider = data$funnel_consider_mr |> setNames(c("jicama", "kohlrabi")),
        buy = data$funnel_buy_mr |> setNames(c("jicama", "kohlrabi"))
    ),
    axes = list(
        tibble::tibble(
            value = c("aware", "consider", "buy"),
            label = c("Aware", "Consider", "Buy"),
            description = NA
        ),
        tibble::tibble(
            value = c( "jicama", "kohlrabi"),
            label = c("Jicama", "Kohlrabi"),
            description = NA
        )
    ),
    label = "Funnel 2D MR"
)

# Can select over either axes
data$funnel_mr_2d[["aware"]]

data$funnel_mr_2d[[list(NULL, "jicama")]]


# Can apply functions to every column
data |>
    group_by(healthy_eater) |>
    summarize(mean(funnel_mr_2d, from = "selected", na.rm = TRUE))
```


### In "lake" format (`variable_format = "lake"`)

The data lake stores categorical data as strings of the category id. To load the data
in this format, use option `variable_format="lake"`.

```{r as_crunch_vars}
cr_read_data(project_dir, as_crunch_variables = FALSE)
```


### Array options
By default, `cr_read_data()` reads array columns as 
["packed" `data.frame` columns](https://tidyr.tidyverse.org/reference/pack.html).
This format is nice because it will never lead to duplicated column names and 
because it allows naturally referring to the array variable columns as a group.

```{r packed_examples}
# The column is itself a `data.frame`
data$enjoy_mr

# So you can refer to parts of it as you would a data.frame
# A single axis item with `$`
data$enjoy_mr$enjoy_mr_savory

# A few items with `[`
data$enjoy_mr[2:3]

# tidyselect syntax
data$enjoy_mr |> select(ends_with("y"))

# The `array_apply` function makes applying a function to all
# axis items easier
standardize <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x)
data |>
    mutate(ratings_std = array_apply(ratings_numa, standardize)) |>
    pull(ratings_std)
```


Option `array_strategy="qualified"` reads the data with as regular columns, and with their names the array name separated
by the subvariable name with a `_`.

```{r array_strategy_qualified}
cr_read_data(project_dir, array_strategy = "qualified") |> 
    dplyr::select(starts_with("enjoy_mr"))

# `enjoy_mr$enjoy_mr_savory` -> `enjoy_mr_enjoy_mr_savory`
```


`array_strategy="unqualified"` reads the columns with just the subvariable names (and will error if there are duplicate
names in across array columns, which you can fix with `name_repair="unique"` if you really want).
```{r array_strategy_unqualified}
cr_read_data(project_dir, array_strategy = "unqualified") |> 
    dplyr::select(starts_with("enjoy_mr"))

# `enjoy_mr$enjoy_mr_savory` -> `enjoy_mr_savory`
```


## Using s3 buckets

The APIs used above to download a dataset to your local computer directly have a relatively
low limit on the size of the data. For larger exports, users can export to an s3 bucket that 
they control.

The S3 bucket must be registered to crunch as a "data destination", like so:
```{r create_destination, eval=FALSE}
destination <- cr_create_destination(
    "gfe-crunch-lake-pkg",
    data = list(type = "s3folder", bucket = "crunch-rteam-prod", base = "lake")
)
```

Each destination must have a unique name (I believe this is per account, not user). Also,
the crunch server must be given access to the s3 bucket if it is not public (unclear
what this process is).

After creating, you can refer to a data destination with `cr_destination()` and provide
it to `cr_export()`.
```{r get_destination, eval=FALSE}
destination <- cr_destination("gfe-crunch-lake-pkg")

s3files <- cr_export(ds, path = "vignette_test", destination = destination)
```

The `arrow` package allows connecting to s3 buckets directly, so it is possible
to read the data directly from the s3 bucket as you would with a local file.
See [`arrow::s3_bucket()`] for details about the ways to connect, and 
`cr_clip_to_env()`, which helps getting credentials from your clipboard.

```{r read_from_destination, eval=FALSE}
cr_read_data(s3files, region = "eu-west-1")
```
