<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Variables: cleaning and defining metadata}
-->

[Previous: create and load datasets](datasets.html)

```{r, results='hide', echo=FALSE, message=FALSE}
## Because the vignette tasks require communicating with a remote host,
## we do all the work ahead of time and save a workspace, which we load here.
## We'll then reference saved objects in that as if we had just retrieved them
## from the server
library(crunch)
load("variables.RData")
```

# Manipulating variables within datasets

Once you've uploaded data to Crunch, you'll likely want to spend some time cleaning up the metadata associated with your dataset. Legacy statistical programs offer only limited support for variable metadata, both in type and quality. Because Crunch facilitates data visualization, collaboration, and sharing results with others, making your dataset presentation-quality is worthwhile, and this often requires additional work after uploading a data file. 

To save yourself work later, strive to have the survey scripted in QSL so that the metadata is automatically included in Crunch via the uploader. This means to utilize `varlabels` and `topics` in QSL wherever possible. If you have questions about this contacts your survey scripter. 

Most of the operations described below can also be accomplished in the web client (GUI). However, users comfortable with scripting may find it faster or easier to automate these actions with the `crunch` library in R. Such edits made within R are carried out on the remote Crunch dataset, thereby keeping data in sync across all client applications.

## A rose by any other moniker: "names" and "aliases"

Crunch takes the principled stand that working with data in the 21st Century should not be constrained by legacies of the punch-card era. Variables should have "names" that are human-readable and searchable by their meaning---there is no reason to constrain variable names to be eight characters, all caps, etc. "Aided awareness: coffee roasters" is much nicer and more presentable than "Q2B_V1".

At the same time, shorter, user-defined, unique identifiers for variables do have their uses. For one, it's what most any legacy statistical software uses for its identifiers, so retaining them on import will help us line up variables when appending a subsequent batch of imported data, for example. For another, when interacting with a dataset from the command line, it can be useful to have shorter, machine-friendlier references. 

So, Crunch stores two user-settable identifiers for variables. What you may have thought of as a variable "label", Crunch elevates to the status of "name". What you may be used to thinking of as a variable "name", Crunch calls "alias".

Aliases and names have slightly different validation constraints. Aliases must be unique across the entire dataset, including among array [subvariables](array-variables.html). "Names", however, only must be unique within [variable order groups](variable-order.html). Any string is valid for either alias or name, though you may want more machine-friendly strings as aliases. In most cases, you probably won't even set aliases, though: they'll be set when you import your dataset and will be whatever the names were in your source data.

In sum, name is crunch alias, label is crunch name.

Except in one place in `crunch`: referencing variables within a dataset.

### Accessing variables within a dataset

When dealing with variables within a dataset, the alias is used to identify variables. This is because (1) aliases are typically what were used to identify variables in whatever format from which the dataset was imported, and consequently, (2) aliases are typically more machine-friendly as names, less likely to contain characters that are not valid as variable names in R.

Because the `names` attribute is used for indexing elements in R, if we want to extract variables based on alias, it means that the `names` attribute of dataset must actually expose aliases. This may be dissonant, but it has some nice properties. For one, comparing our Crunch dataset with the `data.frame` from which it was created, their `names` attribute have the same contents:

```{r, eval=FALSE}
identical(names(ds), names(economist))
```
```
## [1] TRUE
```

You can reference and extract variables from a dataset as if it were a `data.frame`, using the `$`, `[`, and `[[` methods. 

```{r, eval=FALSE}
track.local <- ds$track # remember that we called the dataSET up on Crunch 'ds'
track.local
track.local <- ds[['track']] # the same thing as using the '$'
track.local
```

Note how the `track.local` prints a summary table of the variable, not a long vector of the responses themselves. If this was a data.frame in R, it would have printed to the screen a vector of the actual responses (which the code snippet below shows). Later on we will go over how to get the actual vector of responses for a single (or many) variable(s) in case you need them to do e.g. a recode.
```{r, eval=FALSE}
head(economist$track) # Note how this uses 'economist' not 'ds'. Try '? head' if that's new to you
```

Like datasets, variables have various attributes like `name` and `description` that can be set.

```{r, eval=FALSE}
name(track.local) <- "Direction of country"
description(track.local) <- "In your opinon, is the country going in the right direction, or is it on the wrong track?"
```

Two caveats. First, because we first extracted the variable 'track' from the Crunch dataset and saved it locally as 'track.local' before making edits, the Crunch dataset object does not have the new name and description for this variable ('track'). That is, the name and description we just gave 'track.local' are not up on the Crunch dataset yet. But we can easily fix that.

```{r, eval=FALSE}
name(ds$track) <- name(track.local)
name(ds$track)
```

Alternatively, instead of extracting the variable and making the changes locally, it is more efficient to directly modify `track` up on `ds`, and then `ds` would be up to date automatically.

```{r, eval=FALSE}
name(ds$track) <- "Direction of country"
```

### Addressing the variable catalog

It is not always convenient that the `names` attribute of the dataset actually yields *aliases*. Moreover, if we want to edit the Crunch names of many variables, we need a way of accessing the Crunch metadata more directly. It will be very slow to edit each variable in the dataset individually, referencing them with `$`, because each edit would send a request to the server. Instead, we'd rather bundle those into a single request. To do this, we can access the `variables` attribute of the dataset, which is a "variable catalog":

```{r}
class(variables(ds))
```

In the variable catalog, Crunch names are names, and aliases are aliases. Hence,

```{r}
identical(names(ds), aliases(variables(ds)))
```

but

```{r}
identical(names(ds), names(variables(ds)))
```

because above we changed the name of `ds$track` to "Direction of country" (see the fifth element in the vector below).

```{r, eval=FALSE}
head(names(variables(ds)), 10)
```

These attributes all allow assignment with `<-`. The methods `names` and `aliases` yield character vectors, and they take characters in assignment. Hence, you can use any vectorized string manipulation tools available in R, such as regular expressions, to edit variable names efficiently. You can also just supply a replacement vector, like

```{r, eval=FALSE}
names(variables(ds))[6:9] <- c("Favorability of Edward Snowden", 
                               "Approval of Snowden's Leak",
                               "Support for Prosecution of Snowden",
                               "Penalty for Snowden")
head(names(variables(ds)), 10)
```
```{r, echo=FALSE}
head2
```

## Categorical variables

Many variables in survey data are categorial: respondents have a finite set of answers to the survey question, and the answers are first and foremost of a nominal, not quantitative nature. In R, this data type is represented as a `factor`. The response options, are contained in the factor's "levels" as a character vector. Manipulation of these levels is limited and often challenging.

In Crunch, categorical variables' "categories" are objects with richer metadata. 

```{r}
is.Categorical(track.local)
```
```{r, eval=FALSE}
categories(track.local)
```
```{r, echo=FALSE}
track.cats.before
```

### Category attributes

Categories have `names`, the levels in an R factor; numeric `values` which can be used when interpreting the categorical variable as numeric; and `ids`, which are analogous to the integer values that underly an R factor. 

```{r, eval=FALSE}
names(categories(track.local))
```
```{r, eval=FALSE}
values(categories(track.local))
```
```{r, eval=FALSE}
ids(categories(track.local))
```

Categories also have their own "missing" status. Indeed, because Crunch supports more complex missing values than does R, multiple categories can be marked as missing: there's not a single "NA" value, which can be very handy. We can easily change which categories are marks as missing.

```{r, eval=FALSE}
is.na(categories(track.local)) # is.na() tells us which categories are currently 'missing'
is.na(categories(track.local)) <- 'Not sure' # changes 'Not Sure' to be missing now
is.na(categories(track.local))
is.na(categories(track.local))['Not sure'] <- FALSE # reverses it. Another way to write this
is.na(categories(track.local))
```

You can change categories names and values, but ids cannot be changed: they are immutable references to values within the column of data on the server. 

```{r, eval=FALSE}
names(categories(track.local))[1] <- "Right track" # assigning a category name
categories(track.local)
ids(categories(track.local)) <- sample(ids(categories(track.local)), replace=FALSE)
```

```
## Error : Cannot modify category ids
```

### Reordering categories

Categories can also be reordered by index, like any list object

```{r, eval=FALSE}
categories(track.local) <- categories(track.local)[c(1,3,2)] # we left out 4 too
categories(track.local)
```

As with all other metadata edits discussed, updating with these methods automatically sends the changes to the server, so your local edits are reflected in the cloud.

## Hiding variables

Datasets often contain variables that you may want to *use* -- as a derived variable, a transformation, or a recode -- but that may not be relevant for the analytic tasks at hand. In short, you want to hide them. They aren't deleted, so you can restore them if you need them later, but they no longer clutter the dataset "workspace" and they continue to exist as the seed for whatever dervied variable etc... that you originally wanted the variable for.

You use the `hideVariables` function to tell Crunch which variables you want to hide. You give it a vector of (one or many) characters. The `hiddenVariables` function returns the names (not aliases) of all the hidden variables. To return aliases of hidden variables see the example below.

```{r, eval=FALSE}
hiddenVariables(ds)
hideVariables(ds, "track")
ds <- refresh(ds) # refresh to make sure the variable is hidden
hiddenVariables(ds) # returns the variable *name* not alias
hiddenVariables(ds, 'alias') # Note how returns the alias instead
unhideVariables(ds, 'track')
ds <- refresh(ds) # refresh to make sure the variable is hidden
```

## Deleting variables

Sometimes you do want to delete variables permanently. Doing so is easy, but we have some protections in place to keep you from accidentally deleting things from a dataset that may be shared with many people on the server.

To delete, you can assign `NULL` in to the dataset for that variable, just like you were removing a column from a `data.frame`. In our example dataset, we have two copies of a voter-registration variable, "votereg_new" and "votereg_old". Let's kill the old one:

```{r, eval=FALSE}
ds$votereg_old <- NULL # type 'n' (no) when it prompts you.
```
```
## Really delete "votereg_old"?
```

The delete function requires confirmation when you're running from an interactive session, just to make sure you aren't accidentally assigning something in that is NULL and deleting your variable. If you know that you want to delete the variable, you can give your approval in advance by wrapping it in a `with` statement, using the `consent` context manager:

```{r, eval=FALSE}
with(consent(), ds$votereg_old <- NULL)
"votereg_old" %in% names(ds)
```
```
## [1] FALSE
```

The `with(consent(), ...)` pattern works everywhere in `crunch` that requires confirmation to do an action, such as deleting datasets.

[Next: create and manipulate array variables](array-variables.html)