<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Variable Order}
-->

[Previous: array variables](array-variables.html)

```{r, results='hide', echo=FALSE, message=FALSE, eval=FALSE}
library(rcrunch)
load("variable-order.RData")
```

# Variable Order and Organization

In the web application, variables in a dataset are displayed in a list on the left side of the screen.

<!-- screenshot -->

Typically, when you import a dataset, the variable list is flat, but it can be organized into an accordion-like hierarchy. The variable organizer in the GUI allows you to perform this organization visually, but you can also manage this metadata from `rcrunch`.

## Viewing the order

Picking up with the dataset we used in the Array Variables vignette, we can fetch the `ordering` property of the dataset to view the current organization:

```{r, eval=FALSE}
ordering(ds)
```
```{r, echo=FALSE, eval=FALSE}
ordering(start)
```

## Group names

## Creating groups

## Reordering groups and entities

## Nested groups

[Next: transforming and deriving](derive.html)